Supplement to Human Endogenous Retroviral Transcripts Show Potential as Novel, Tumor Specific Biomarkers for Colorectal Cancer

### 2020-12-31 manuscript/main.tex
forging ahead through discussion
fix final results section
finish text, redo figures

    \title{Potential Novel Immunotarget Revealed in Mega Analysis of Colorectal Transcriptomes}

@@

\newpage
\section*{Acknowledgements}

People we want to thank. Grant support.

@@

\begin{figure}[ht]
  \makebox[\textwidth][c]
  {
    \includegraphics[width=0.9\textwidth]{fig1}
  }
  \caption{Summary of unannotated human endogenous retroviral sequences}
  \label{fig:pre_sum}
\end{figure}

\begin{figure}[ht]
  \makebox[\textwidth][c]
  {
    \includegraphics[width=0.9\textwidth]{fig2}
  }
  \caption{Differential expression results for unannotated HERV genes}
  \label{fig:dge_res}
\end{figure}

\begin{figure}[ht]
  \makebox[\textwidth][c]
  {
    \includegraphics[width=0.9\textwidth]{fig3}
  }
  \caption{Global estimates of HERV expression across tissue phenotypes}
  \label{fig:heat}
\end{figure}

>>>>

    %\title{Potential Novel Immunotarget Revealed in Mega Analysis of Colorectal Transcriptomes}
    \title{Human Endogenous Retroviral Transcripts Show Potential as Novel, Tumor Specific Biomarkers for Colorectal Cancer}

@@

\newpage
\section*{Acknowledgements}
\input{secs/acknowledge.tex}

@@

\begin{figure}[ht]
  \makebox[\textwidth][c]
  {
    \includegraphics[width=0.9\textwidth]{fig1}
  }
  \caption{Summary of unannotated human endogenous retroviral sequences found in colorectal transcriptomes}
  \label{fig:pre_sum}
\end{figure}

\begin{figure}[ht]
  \makebox[\textwidth][c]
  {
    \includegraphics[width=0.9\textwidth]{fig2}
  }
  \caption{Differential expression of unannotated human endogenous retroviral loci}
  \label{fig:dge_res}
\end{figure}

\begin{figure}[ht]
  \makebox[\textwidth][c]
  {
    \includegraphics[width=0.9\textwidth]{fig3}
  }
  \caption{Global estimates of proviral expression across tissue phenotypes}
  \label{fig:heat}
\end{figure}

### 2020-12-31 manuscript/secs/abstract.tex
finish text, redo figures

\textbf{Background}:
Microsatellite unstable, mismatch repair deficient metastatic cancers can be treated with immune checkpoint inhibitors, but new therapeutic approaches are needed to extend the benefits of immune modulators to the majority of patients with microsatellite stable tumors.
Cancer specific over expression of human endogenous retroviral elements could be exploited to enhance immune infiltration of microsatellite stable colorectal cancer and render these tumors more susceptible to immune checkpoint inhibition.
To identify novel targets for immune mediated cancer therapies, we tested for cancer specific over expression of endogenous retroviral transcripts in a large cohort of colorectal transcriptomes.

\textbf{Methods}:
A recently curated RNA sequencing dataset composed of 1,139 human primary colorectal tissue samples divided into three cohorts was re-processed using a novel annotation of repeat element transcripts to quantify levels of repeat element expression in colorectal transcriptomes.
A total of NUMBER healthy, NUMBER tumor-adjacent normal, and NUMBER tumor samples were profiled with short read sequencing.
Transcripts were quantified with Salmon using the GENCODE v36 transcriptome augmented to include unannotated exons from long terminal repeat elements classified as human endogenous retroviruses (HERVs) in the Dfam database.
Associations between transcript expression and tissue phenotype were tested using DESeq2 in the R statistical language.

\textbf{Results}:
Out of 5,025 unannotated HERV genes measured in 834 samples, including NUMBER healthy, NUMBER tumor-adjacent normal, and NUMBER tumor samples, 150 were expressed within one standard deviation of median protein coding gene expression in at least one tissue phenotype.
Over half of the 150 HERV genes with moderate expression were in the H (26) or K (51) classes.
Tumor specific over expression was observed in 11 of the 150 HERV genes tested.

\textbf{Conclusions}: The conclusions go here.

>>>>

\textbf{Background}:
Colorectal cancer is the second leading cause of cancer death in the United States.
Identification of novel genomic features with potential utility as diagnostic or prognostic biomarkers or therapeutic targets could help improve detection and treatment of patients with colorectal cancer.
Features associated with immune modulation are of particular relevance given the emergence of immune checkpoint inhibition as an effective treatment for advanced disease.
Endogenous retroviral elements expressed in tumors could be exploited for diagnostic or therapeutic use.
To identify novel markers and targets for immune mediated cancer therapies, we tested for cancer specific over expression of non-canonical endogenous retroviral transcripts in a large cohort of colorectal transcriptomes.

\textbf{Methods}:
A curated RNA sequencing dataset composed of 1,139 human primary colorectal tissue samples divided into three independent cohorts was re-processed using a novel annotation of repeat element transcripts to quantify levels of human endogenous retroviral mRNA in colorectal transcriptomes.
A total of 474 healthy, 133 tumor adjacent normal, and 532 tumor samples were profiled with short read sequencing.
Transcripts were quantified with Salmon using the GENCODE v36 transcriptome augmented to include unannotated exons from long terminal repeat elements classified as human endogenous retroviruses in the Dfam database.
Associations between transcript expression and tissue phenotype were tested using \emph{DESeq2} in the R statistical programming language.

\textbf{Results}:
Out of 5,025 unannotated proviral loci queried in our discovery analysis, 150 were expressed within one standard deviation of median protein coding gene expression in at least one tissue phenotype.
Over half of the 150 loci were from two groups of endogenous retrovirus (51 from K, 26 from H).
Tumor specific over expression was observed in 11 of the 150 loci tested (\emph{P} adjusted < 0.05), and seven were validated in an independent cohort.
Widespread proviral expression was also found in healthy tissues.

\textbf{Conclusions}:
In the largest unbiased assessment of non-canonical human endogenous retroviral transcript expression in colorectal transcriptomes, we found relatively few loci expressed at levels similar to canonical protein coding genes.
Nevertheless, we identified seven proviruses with tumor specific expression patterns.
These proviruses have potential as novel biomarkers or immunotargets and should be prioritized in future investigations.

### 2020-12-31 manuscript/secs/discussion.tex
finish text, redo figures

In this study, we performed the largest unbiased assessment of transcribed but unannotated human endogenous retroviral elements in colorectal tissue to date.
We found several unique proviral loci with tumor specific over expression and many other loci with relatively high expression in healthy samples.
Human endogenous retroviral elements are normally transcriptionally repressed repetitive sequences that are often de-repressed in epigenetically altered cancer cells \citep{Babaian2016}.
HERV transcripts have been shown to engage pattern recognition receptors and activate interferon responses in cancer cell lines \citep{Chiappinelli2015, Roulois2015} and have been associated with tumor infiltration by T regulatory cells \citep{Desai2017}, response to PD-L1 blockade in urothelial carcinoma, and survival in subjects with CRC \citep{Solovyov2018}.
Our results are important because the tumor specific HERV sequences we identified could be used to distinguish tumor from non-tumor tissue for screening and surveillance and to predict or even modify response to immunotherapies.

Three of the seven HERV proviruses we validated as tumor specific were from group H.
Over expression of HERV-H proviruses in CRC has been demonstrated previously with microarray \citep{Perot2012} and quantitative real time polymerase chain reaction \citep{Perot2015}.
The prominence of HERV-H transcripts in RNA-seq from CRC samples of The Cancer Genome Atlas, from which many of the samples in the current study were generated, has also been previously reported \citep{Desai2017, Attig2019}.
Therefore, our findings are generally consistent with prior work.
However, we provide new evidence of the reliability of tumor specific over expression by jointly analyzing a large number of samples from different studies.
Furthermore, our analysis was not restricted to a particular group of proviruses.
The unbiased nature of our assessment provides important context for the community of researchers selecting genomic features to develop as biomarkers and immunotargets for CRC.

@@

This study also demonstrated the surprising frequency of HERV expression in healthy colorectal tissue.
Interpretation of this result is challenging.
HERVs have been shown to have rare functional roles in healthy cells \citep{Rote2004} and could conceivably have a role in maintaining healthy colorectal epithelium, but this is speculative.
Over expression in our healthy samples could also result from technical (i.e. sequencing protocols) or demographic (i.e. age) differences between phenotypes.
Latent factor analysis is expected to adjust for multiple confounders, including subject age and sequencing artifacts, but the adjustment may be insufficient.
Although surprising, the detection of high levels of HERV expression in healthy samples reduced the likelihood that using a reference HERV transcriptome assembled exclusively from contigs of cancer cells had biased our results.

The major limitation of this study was its reliance on complex bioinformatic and statistical methods to aggregate heterogeneous RNA-seq datasets and quantify poorly annotated transcripts.
Specifically, the definition and quantification of HERV elements was of critical importance.
We used a published, hidden Markov model based annotation of a \emph{de novo} assembled cancer transcriptome to define and locate the HERV transcripts and loci we studied \citep{Attig2019}.
We used a published, alignment-free, k-mer matching quantification algorithm to probabilistically estimate transcript abundance of repetitive sequences that may otherwise have been excluded from analysis due to multiple possible mappings \citep{Patro2017}.
We used state of the art statistical packages designed specifically for high dimensional genomics analyses to perform association tests \citep{Love2014, sva}.
We also pursued multiple validations of our findings and methods to ensure reproducibility.

The aim of this work was to prioritize HERV transcripts for potential development as novel biomarkers or immunotargets in the management of CRC.
Markers and targets for immune modulation are of particular relevance given the emergence of immune checkpoint inhibition as an effective treatment for advanced CRC.
The seven proviruses we identified were consistently and reliably over expression in colorectal tumors and warrant future investigation.

>>>>

In this study, we performed the largest unbiased assessment of transcribed but unannotated human endogenous retroviral elements in colorectal tissue to date.
We found several unique proviral loci with tumor specific over expression and many other loci with relatively high expression in healthy samples.
Human endogenous retroviral elements are normally transcriptionally repressed repetitive sequences that are often de-repressed in epigenetically altered cancer cells \citep{Babaian2016}.
HERV transcripts have been shown to engage pattern recognition receptors and activate interferon responses in cancer cell lines \citep{Chiappinelli2015, Roulois2015} and have been associated with tumor infiltration by T regulatory cells \citep{Desai2017}, response to immune checkpoint blockade in urothelial carcinoma, and survival in subjects with CRC \citep{Solovyov2018}.
Our results are important because the tumor specific HERV sequences we identified could be used to distinguish tumor from non-tumor tissue for screening and surveillance and to predict or even modify response to immunotherapies.

Three of the seven HERV proviruses we validated as tumor specific were from group H.
Over expression of HERV-H proviruses in CRC has been demonstrated previously with microarray \citep{Perot2012} and quantitative real time polymerase chain reaction \citep{Perot2015}.
The prominence of HERV-H transcripts in RNA-seq from TCGA CRC samples has also been previously reported \citep{Desai2017, Attig2019}.
Therefore, our findings are generally consistent with prior work.
However, we provide new evidence of the reliability of tumor specific over expression by jointly analyzing a large number of samples from different studies.
Furthermore, our analysis was not restricted to a particular group of proviruses.
The unbiased nature of our assessment provides important context for the community of researchers selecting genomic features to develop as biomarkers and immunotargets for CRC.

@@

This study also demonstrated widespread expression of a subset of HERVs in healthy colorectal tissue.
Interpretation of this result is challenging.
HERVs have been shown to have rare functional roles in healthy cells \citep{Rote2004} and could conceivably have a role in maintaining healthy colorectal epithelium, but this is speculative.
Over expression in our healthy samples could also result from technical (i.e. sequencing protocols) or demographic (i.e. age) differences between phenotypes.
Latent factor analysis is expected to adjust for multiple confounders, including subject age and sequencing artifacts, but the adjustment may be insufficient.
Although surprising, the detection of high levels of HERV expression in healthy samples reduced the likelihood that using a reference HERV transcriptome assembled exclusively from contigs of cancer cells had biased our results.

The major limitation of this study was its reliance on heterogeneous RNA-seq datasets and quantifications of short reads of repetitive sequences that are infamously challenging to estimate accurately.
To address these limitations, we built a complex bioinformatic pipeline based on state of the art software.
Specifically, our definition of HERV elements was of critical importance.
We used a published, hidden Markov model based annotation of a \emph{de novo} assembled cancer transcriptome to define and locate the HERV transcripts and loci we studied \citep{Attig2019}.
We used a published, alignment-free, k-mer matching quantification algorithm to probabilistically estimate transcript abundance of repetitive sequences that may otherwise have been excluded from analysis due to multiple possible mappings \citep{Patro2017}.
We used actively developed and updated statistical packages designed specifically for high dimensional genomics analyses to perform association tests \citep{Love2014, sva}.
We also pursued multiple validations of our findings and methods to ensure reproducibility.

The aim of this work was to prioritize HERV transcripts for potential development as novel biomarkers or immunotargets in the management of CRC.
Markers and targets for immune modulation are of particular relevance given the emergence of immune checkpoint inhibition as an effective treatment for advanced disease.
The seven proviruses we identified were consistently and reliably over expressed in colorectal tumors and warrant future investigation.

### 2020-12-31 manuscript/secs/introduction.tex
finish text, redo figures

Screening and treatment strategies have begun to rely on genomic biomarkers for prediction and prognosis, but novel genomic features with better sensitivity and specificity are under active investigation \citep{Imperiale2014, Sveen2020}.
Among emerging treatment strategies, immune checkpoint inhibition is effective against metastatic tumors with mismatch repair defficiency and high microsatellite instability \citep{Le2015, Overman2018}.
Unfortunately, only 4\% of metastatic CRCs display high microsatellite instability \citep{Ganesh2019}.
Identification of additional classes of tumor with potential susceptibilty to immunotherapy and development of treatments to improve tumor immunogenicity are required to help extend the benefit of checkpoint blockade.
Endogenous retroviral elements expressed in cancers may offer a solution.

Human endogenous retroviruses (HERVs) are proviral elements integrated into the human genome under conditions that render them incompetent of replication or infection.
Residual HERV elements in contemporary human genomes are the consequence of retroviral infections of germ cells millions of years ago.
Since then, and like the rest of the human genome, HERV elements have accumulated point mutations and segmental deletions.
These alterations in HERV sequences disrupt assembly of functional viral machinery.
Nevertheless, some HERV sequences can still be transcribed and translated, with some protein products even processed as endogenous antigens and presented on major histocompatibility complex class I molecules \citep{Boller1997, Rooney2015}.
In healthy tissues, DNA methylation and histone modification can silence transcription of HERV proviruses, but epigenetic modifications in the context of oncogenesis can de-repress HERV transcription and occasionally lead to translation of neoantigens \citep{Menendez2004, Wiesner2015}.

The HERV transcriptome is a promising source of biomarkers for cancer detection and treatment because of the putative cancer specificity of HERV mRNA expression.
A major challenge to the development of HERVs as cancer biomarkers is the absence of consensus on how to define and identify HERV elements.
Proviruses are generally absent from human gene and transcript annotations, and their structural features make them hard to identify in agnostic surveys based on DNA or RNA sequencing \citep{ERVmap2018, Treangen2011}.
Proviral genomes are flanked by repetitive elements that often cannot be mapped with short read sequencing.
Furthermore, HERVs are transposable elements that have generated many similar proviral integrations throughout the human genome.
The multitude of similar sequences across the genome impairs accurate high throughout read alignment.
As a result, the genomic locations and transcriptional abundances of most HERVs are not known with confidence.

Recently, new approaches have been developed to overcome traditional limitations in HERV annotation and quantification \citep{Attig2017, Attig2019, ERVmap2018, Telescope2019, Grabski2020}.
Here, we extend advances in the characterization of HERV transcripts to discover novel CRC specific expression patterns in human colorectal transcriptomes.
We hypothesized that HERV proviral elements would be preferentially transcribed in tumor samples compared to healthy and tumor adjacent, histologically normal samples.
We tested this hypothesis in a large cohort of healthy, tumor adjacent, and tumor RNA sequencing data and found several HERV proviruses over-expressed in tumors but many that had greater expression in healthy tissue.

>>>>

Screening and treatment strategies have begun to rely on genomic biomarkers for prediction and prognosis, and novel genomic features with better sensitivity and specificity are under active investigation \citep{Imperiale2014, Sveen2020}.
Among emerging treatment strategies, immune checkpoint inhibition is effective against metastatic tumors with mismatch repair defficiency and high microsatellite instability \citep{Le2015, Overman2018}.
Unfortunately, only 4\% of metastatic CRCs display high microsatellite instability \citep{Ganesh2019}.
Identification of additional classes of tumor with potential susceptibilty to immunotherapy and development of treatments to improve tumor immunogenicity are required to help extend the benefit of checkpoint blockade.
Endogenous retroviral elements expressed in cancers may offer a solution.

Human endogenous retroviruses (HERVs) are formerly transposable, proviral elements previously integrated into the human genome and now incompetent of transposition, replication, or infection \citep{Babaian2016}.
Residual HERV elements in contemporary human genomes are the consequence of retroviral infections of germ cells millions of years ago.
Since then, like the rest of the human genome, HERV elements have accumulated point mutations and segmental deletions.
These alterations in HERV sequences disrupt assembly of functional viral machinery.
Nevertheless, some HERV sequences can be transcribed and translated, and some protein products can be processed as endogenous antigens and presented on major histocompatibility complex class I molecules \citep{Boller1997, Mullins2012, Rooney2015}.
In healthy tissues, DNA methylation and histone modification silence HERV transcription, but epigenetic modifications in the context of cancer can de-repress HERV transcription and occasionally lead to translation of neoantigens \citep{Babaian2016, Menendez2004, Wiesner2015}.

The HERV transcriptome is a promising source of novel genomic features for cancer detection and treatment because of the putative cancer specificity of HERV mRNA expression and its association with immune modulation \citep{Chiappinelli2015, Roulois2015, Desai2017, Solovyov2018}.
A major challenge to the development of HERVs as cancer biomarkers is the absence of consensus on how to define and identify HERV elements.
Proviruses are generally absent from human gene and transcript annotations, and their structural features make them hard to identify in agnostic surveys based on DNA or RNA sequencing \citep{ERVmap2018, Treangen2011}.
Proviral genomes are flanked by repetitive elements that often cannot be mapped with short read sequencing.
Furthermore, HERVs were originally transposable elements that have generated many similar proviral integrations throughout the human genome.
The multitude of similar sequences across the genome impairs accurate high throughout read alignment.
As a result, the genomic locations and transcriptional abundances of most HERVs are not known with confidence.

Recently, new approaches have been developed to overcome traditional limitations in HERV annotation and quantification \citep{Attig2017, Attig2019, ERVmap2018, Telescope2019, Grabski2020}.
Here, we extend advances in the characterization of HERV transcripts to discover novel CRC specific expression patterns in human colorectal transcriptomes.
We hypothesized that HERV proviral elements would be preferentially transcribed in tumor samples compared to healthy and tumor adjacent, histologically normal samples.
We tested this hypothesis in a curated dataset of healthy, tumor adjacent, and tumor RNA sequencing (RNA-seq) reads and found several HERV proviruses over expressed in tumors but many that had greater expression in healthy tissue.

### 2020-12-31 manuscript/secs/methods.tex
finish text, redo figures

\subsection*{Sample Selection}
Bulk RNA sequencing (RNA-seq) from a recently curated set of 1,139 flash frozen colorectal tissue samples was re-processed and analyzed for the presence of unannotated HERV genes \citep{Dampier2020}.
For discovery of associations between tissue phenotypes and HERV expression, paired end sequencing from 834 colon biopsies was used.
Of those biopsies, 462 were healthy tissue, 61 were tumor adjacent normal tissue, and 311 were tumors.
These samples are referred to as the discovery cohort.

For validation of discovered associations, two independent cohorts were used.
The first (independent cohort \#1) included single end sequencing from 275 colon biopsies of which NUMBER were healthy, NUMBER were tumor adjacent, and NUMBER were tumor.
The second (independent cohort \#2) included paired end sequencing from 30 pairs of tumor and matched adjacent normal biopsies (i.e. 15 samples of both tissue phenotypes paired by subject).
An important characteristic of the discovery cohort and independent cohort \#1 was the independence of all tumor and tumor adjacent samples.
To permit valid inference from regular linear models, only one tissue phenotype was used per subject.

Samples were gathered from publicly available RNA-seq datasets hosted on the Genomic Data Commons and Sequence Read Archive as well as from a recently generated dataset of healthy mucosal biopsies obtained during screening colonoscopies at the Catalan Institute of Oncology in Barcelona, Spain \citep{Dampier2020, DiezObrero2020}.
A complete description of sample selection and cohort construction for the curated dataset is available in Dampier et al. \citep{Dampier2020}.
RNA-seq reads for all samples were stored and analyzed in FASTQ format.
For samples downloaded in BAM format, Biobambam2 v2.0.87 \citep{Tischler2014} was used to convert reads to FASTQ format.

\subsection*{Human Endogenous Retrovirus Gene Annotation}
HERVs were identified using the genomic coordinates (GRCh38) and feature names of exons and spliced transcripts of HERV elements from a recently published human cancer transcriptome annotated with the GENCODE v24 basic gene model \citep{Frankish2018} and a model of genome-wide repeat elements \citep{Attig2019}.
The cancer transcriptome was assembled from RNA-seq reads of 24 subjects from each of 32 cancer types obtained from The Cancer Genome Atlas (TCGA), as previously described \citep{Attig2019}.
Names and coordinates of repeat elements were generated using RepeatMasker \citep{Smit2015} configured with nhmmer \citep{Wheeler2013} in sensitive mode using the Dfam 2.0 library (v150923) \citep{Hubley2015}, as previously described \citep{Attig2017}.
The method relies on profile hidden Markov models representing known human repeat families to classify repeat elements throughout the genome.

Previously unannotated HERV genes were selected for the current study in multiple steps.
A complete gene transfer format (GTF) \citep{GTF} annotation of the repeat element inclusive cancer transcriptome was downloaded from Attig et al. \citep{Attig2019}.
The annotation was parsed with a custom Python script to produce a new annotation in BED12 \citep{BED12} format.
Transcripts were selected based on three criteria: i) overlap of at least one exon with a long terminal repeat element, ii) classification of at least one exon as HERV derived, and iii) abscence of any exon overlapping an annotated GENCODE gene.
All transcripts labeled with the same repeat element identifier (i.e. repeat element class along with genomic coordinates) were considered isoforms of the same HERV gene.
Genomic coordinates were extracted for all constituent exons of selected transcripts to permit inference of splice sites and exclusion of intronic sequences.
For display purposes, repeat element identifier genomic coordinates were converted to cytobands.

\subsection*{Transcript Quantification}
Transcript quantification was performed with Salmon v1.2.1 \citep{Patro2017} in mapping-based mode with the \verb|--gcBias| and \verb|--validateMappings| flags set.
Inputs to the Salmon \verb|quant| command were RNA-seq reads in FASTQ format and the Salmon index.
To create the Salmon index, a custom reference transcriptome was built from a list of GENCODE v34  transcript sequences \citep{Frankish2018} and the list of unannotated HERV transcript sequences.
GENCODE sequences were downloaded from the GENCODE FTP site \citep{GENCODE-transcripts}.
HERV sequences were extracted from the GRCh38 reference genome \citep{ENCODE-GRCh38} using the BEDTools \citep{Quinlan2010} \verb|getfasta| command with the \verb|-split| option and the custom BED12 annotation of selected HERV transcripts.
Following the recommendation of Salmon's authors \citep{SalmonDecoys}, contig sequences from the GRCh38 reference genome were subsequently added to the target transcript sequences as decoy sequences to generate a decoy-aware transcriptome for indexing.

\subsection*{Gene Expression Analysis}
Associations between HERV gene expression and tissue phenotype were tested using the \emph{Bioconductor} suite of genomics analysis packages \citep{bioc} in the R statistical programming language \citep{R}.
Specifically, \emph{tximport} \citep{Soneson2015} was used to extract transcript length and abundance estimates from Salmon and summarize them to the gene level for input into \emph{DESeq2} \citep{Love2014}.
Count matrices of HERV and GENCODE genes were prefiltered to select genes with a count of at least one read in at least half of the samples in each cohort.
Factors to normalize counts for sequencing depth and transcript length were calculated with \emph{DESeq2}.
Latent factors of non-biological variation were estimated with \emph{sva} \citep{sva}.
Genes of interest for association testing were selected in several steps.

First, counts of all genes were transformed with the variance stabilizing transformation of \emph{DESeq2} and adjusted for batch effects with \emph{limma} \citep{Ritchie2015}.
Second, adjusted counts of all annotated genes were filtered to retain only protein coding genes.
Third, median and standard deviation of protein coding gene expression were calculated separately for each tissue phenotype in each cohort.
Finally, adjusted counts of HERV genes were filtered to retain only HERVs expressed within a single standard deviation of median protein coding gene expression in at least one tissue phenotype.

Counts of all protein coding genes and HERVs with sufficient expression were modeled with phenotype and latent factors as explanatory variables using \emph{DESeq2}.
Significant associations between gene count and tissue phenotype were identified by testing whether the absolute value of the effect of phenotype was at least two fold change for a given gene.
Multiple hypothesis testing correction was performed using a false discovery rate of 5\%.

>>>>

\subsection*{Sample selection}
Bulk RNA-seq from a curated set of 1,139 flash frozen colorectal tissue samples was re-processed and analyzed for the presence of unannotated HERV transcripts \citep{Dampier2020}.
For discovery of associations between tissue phenotypes and HERV expression, paired end sequencing from 833 colon biopsies was used.
Of those biopsies, 462 were from healthy, 62 were from tumor adjacent normal, and 309 were from tumor tissue.
This set of samples is referred to as the discovery cohort.

For validation of discovered associations, two independent cohorts were used.
The first (independent cohort \#1) included single end sequencing from 276 colon biopsies of which 12 were healthy, 56 were tumor adjacent, and 208 were tumor.
The second (independent cohort \#2) included paired end sequencing from 30 pairs of tumor and matched adjacent normal biopsies (i.e. 15 samples of both tissue phenotypes paired by subject).
An important characteristic of the discovery cohort and independent cohort \#1 was the independence of all tumor and tumor adjacent samples.
To permit valid inference from regular linear models, only one tissue phenotype was used per subject.

Samples were originally gathered from publicly available RNA-seq datasets hosted on the Genomic Data Commons and Sequence Read Archive as well as from a recently generated dataset of healthy mucosal biopsies obtained during screening colonoscopies at the Catalan Institute of Oncology in Barcelona, Spain \citep{Dampier2020, DiezObrero2020}.
A complete description of sample selection and cohort construction for the curated dataset is available in Dampier et al. \citep{Dampier2020}.
RNA-seq reads for all samples were stored and analyzed in FASTQ format.
For samples downloaded in BAM format, Biobambam2 v2.0.87 \citep{Tischler2014} was used to convert reads to FASTQ format.

\subsection*{Human endogenous retrovirus annotation}
HERVs were identified using the genomic coordinates (GRCh38) and feature names of exons and spliced transcripts of HERV elements from a recently published human cancer transcriptome annotated with the GENCODE v24 basic gene model \citep{Frankish2018} and a model of genome-wide repeat elements \citep{Attig2019}.
The cancer transcriptome was assembled from RNA-seq reads of 24 subjects from each of 32 cancer types obtained from The Cancer Genome Atlas (TCGA), as previously described \citep{Attig2019}.
Names and coordinates of repeat elements were generated using RepeatMasker \citep{Smit2015} configured with nhmmer \citep{Wheeler2013} in sensitive mode using the Dfam 2.0 library (v150923) \citep{Hubley2015}, as previously described \citep{Attig2017}.
The method relies on profile hidden Markov models representing known human repeat families to classify repeat elements throughout the genome.

Previously unannotated HERV genes were selected for the current study in multiple steps.
A complete gene transfer format (GTF) \citep{GTF} annotation of the repeat element inclusive cancer transcriptome was downloaded from Attig et al. \citep{Attig2019}.
The annotation was parsed with a custom Python script to produce a new annotation in BED12 \citep{BED12} format.
Transcripts were selected based on three criteria: i) overlap of at least one exon with a long terminal repeat element, ii) classification of at least one exon as HERV derived, and iii) abscence of any exon overlapping an annotated GENCODE gene.
All transcripts labeled with the same repeat element identifier (i.e. repeat element class along with genomic coordinates) were considered isoforms of the same HERV locus.
Genomic coordinates were extracted for all constituent exons of selected transcripts to permit inference of splice sites and exclusion of intronic sequences.
For display purposes, repeat element identifier genomic coordinates were converted to cytobands based on the cytoBand table downloaded from the UCSC Genome Browser \citep{Furey2003, Navarro2020}.

\subsection*{Transcript quantification}
Transcript quantification was performed with Salmon v1.2.1 \citep{Patro2017} in mapping-based mode with the \verb|--gcBias| and \verb|--validateMappings| flags set.
Inputs to the Salmon \verb|quant| command were RNA-seq reads in FASTQ format and a custom Salmon index.
To create the Salmon index, a custom reference transcriptome was built from a list of GENCODE v34  transcript sequences \citep{Frankish2018} and the list of selected HERV transcript sequences.
GENCODE sequences were downloaded from the GENCODE FTP site \citep{GENCODE-transcripts}.
HERV sequences were extracted from the GRCh38 reference genome \citep{ENCODE-GRCh38} using the BEDTools \verb|getfasta| command \citep{Quinlan2010} with the \verb|-split| option and the custom BED12 annotation of selected HERV transcripts mentioned above.
Following the recommendation of Salmon's authors \citep{SalmonDecoys}, contig sequences from the GRCh38 reference genome were subsequently added to the target transcript sequences as decoy sequences to generate a decoy-aware transcriptome for indexing.

\subsection*{Proviral locus and protein coding gene expression analysis}
Associations between HERV locus expression and tissue phenotype were tested using the \emph{Bioconductor} suite of genomics analysis packages \citep{bioc} in the R statistical programming language \citep{R}.
Specifically, \emph{tximport} \citep{Soneson2015} was used to extract transcript length and abundance estimates from Salmon output and summarize them by HERV locus and GENCODE gene for input into \emph{DESeq2} \citep{Love2014}.
Count matrices of HERV loci and GENCODE genes were prefiltered to select loci and genes with a count of at least one read in at least half of samples in each cohort.
Factors to normalize counts for sequencing depth and transcript length were calculated with \emph{DESeq2}.
Latent factors of non-biological variation were estimated with \emph{sva} \citep{sva} as described in the Supplementary Methods.
Loci and genes of interest for association testing were selected in several steps.

First, counts of all loci and genes were transformed with the variance stabilizing transformation of \emph{DESeq2} and adjusted for batch effects with \emph{limma} \citep{Ritchie2015}.
Second, adjusted counts of all annotated genes were filtered to retain only protein coding genes.
Third, the median and standard deviation of protein coding gene expression were calculated separately for each tissue phenotype in each cohort.
Finally, adjusted counts of HERV loci were filtered to retain only HERVs expressed within a single standard deviation of median protein coding gene expression in at least one tissue phenotype.

Counts of all HERV loci and protein coding genes with sufficient expression were modeled with phenotype and latent factors as explanatory variables using \emph{DESeq2}.
Significant associations between gene count and tissue phenotype were identified by testing whether the absolute value of the effect of phenotype was at least two fold change for a given gene.
Multiple hypothesis testing correction was performed using a false discovery rate of 5\%.

All figures were produced with \emph{ggplot2} \citep{Wickham2016} as described in the Supplementary Methods.
